### 2023-07-28 python_excel_ite.tex
final proofing through submodule1

\ABSTRACT{%
A staple of many spreadsheet based management science textbooks and courses are the use of Excel for activities such as model building, sensitivity analysis, goal seeking and Monte-Carlo simulation. What might those things look like if done with Python? We describe a teaching module in which Python is used to do typical Excel based modeling and data wrangling tasks. In addition, students are exposed to basic software engineering principles including project folder structures, version control, object oriented programming and other more advanced Python skills, creating deployable packages and documentation. The module is supported with Jupyter notebooks, Python scripts, course web pages which include numerous screencasts, and a few GitHub repositories. All of the supporting materials are permissively licensed and freely accessible.
}%

@@

Spreadsheet based modeling has transformed the teaching of management science in business schools over the past twenty plus years. The late 1990's saw first editions of game changing textbooks \cite{winstonPracticalManagementScience2018,ragsdaleSpreadsheetModelingDecision2017} focused on the use of spreadsheets in teaching modeling and management science.  A series of conferences focused on teaching management science with spreadsheets started at Dartmouth in 1998 and several presentations from that conference made up the bulk of papers in the first two issues of the INFORMS Transactions on Education (ITE) journal. The next few decades saw numerous textbooks \cite{powellBusinessAnalyticsArt2019, camm2020business} and teaching papers in journals such as \href{https://pubsonline.informs.org/journal/ited}{ITE}that prominently featured spreadsheet based modeling.

Despite numerous calls for, and predictions of, the eradication of spreadsheets from the business world, Excel and other spreadsheets are \href{https://benn.substack.com/p/the-next-billion-programmers}{alive and well}. Yes, spreadsheets have well known limitations. Spreadsheets are not databases and probably should not be used as one, especially for relational data models. There are better data visualization and exploration tools. There are better dashboarding tools that are more easily maintained in a production environment. Spreadsheet errors can be a \href{https://eusprig.org/research-info/horror-stories/}{huge problem}. Version control is difficult as both data and logic live in the same spreadsheet document. Documentation of multi-step manual data manipulation processes present challenges, it it is done at all -- hindering reproducibility and automation. There are data size limitations and external data links present their \href{https://www.thespreadsheetguru.com/find-remove-external-links/}{own set of challenges}. But, Excel continues to be used for a wide range of quantitative analysis related activities throughout the business world. One particular use is the building and exercising of models of business problems
to explore the impact of important model inputs on different output metrics. This ``what if?'' type modeling is facilitated by Excel's flexibility, numerous built in functions, integrated sensitivity analysis via \href{https://support.microsoft.com/en-us/office/calculate-multiple-results-by-using-a-data-table-e95e2487-6ca6-4413-ad12-77542a5ea50b}{Data Tables},  \href{https://support.microsoft.com/en-us/office/use-goal-seek-to-find-the-result-you-want-by-adjusting-an-input-value-320cb99e-f4a4-417f-b1c3-4f369d6e66c7}{Goal
	Seek} or break-even analysis and visualization capabilities.

@@

At our institution, The EwP module is part of a business course entitled  \href{http://www.sba.oakland.edu/faculty/isken/courses/aap}{Advanced Analytics with Python} (AAP). The students taking this course have already taken the PCDA course within which they learned fundamental Python programming within the context of data analytics over seven weeks. The PCDA course ends with intro to the \href{https://scikit-learn.org/stable/index.html}{scikit-learn library} for doing predictive modeling in Python and the AAP course begins where the PCDA course ends. The EwP module could also be used as part of a semester long Python based analytics course. Most AAP students have also already taken   \href{http://www.sba.oakland.edu/faculty/isken/courses/ba}{Business Analytics} (BA) which is a spreadsheet based introduction to business analytics. We do not strictly require the spreadsheet based course be taken before the PCDA or AAP courses as that course draws students from outside of our business programs who are strictly interested in programmatic tools such as Python. The BA and PCDA courses have historically been offered in face to face mode in a computer teaching lab. Since 2020, these two courses have been also offered as online, asynchronous courses. The newer AAP course has been offered since 2021 in an online asynchronous mode. All three courses have extensive course websites that are publicly accessible, including access to all of the video content and supporting files.

@@

\begin{itemize}
	\item variables, arithmetic and boolean operators, and basic data types,
	\item data structures such as lists, dictionaries, tuples, NumPy arrays and pandas dataframes,
	\item flow control such as branching with \texttt{if \ldots else} and looping with \texttt{for} and \texttt{while}
	\item module imports,
	\item using built in functions and accessing methods and properties of objects,
	\item creating functions,
	\item basic use of NumPy and pandas for data wrangling and analysis,
	\item basic plotting with matplotlib.
\end{itemize}

@@

All of the Python modules in the PCDA course as well as the entire AAP course are taught using a combination of Jupyter notebooks, Python scripts, and videos walking through the notebooks. Jupyter notebooks, part of \href{https://jupyter.org/}{Project Jupyter}, are an example of \textit{literate computing} \cite{perez2015project} and allow one to create rich interactive documents that support iterative exploration and development.

Central to Jupyter notebooks is a mix of \textit{code cells}, \textit{markdown cells} and output cells. Code cells contain executable code while markdown cells contain plain text using \href{URL}{markdown} for styling text. For example, Figure \ref{fig:cells_edit_mode} shows two markdown cells separated by a code cell as they look when in editing mode. After running the cells, they appear as show in Figure \ref{fig:cells_after_running}. After executing a code cell, the output appears directly below it in an output cell. Mixing code and markdown cells makes it quite easy to create interactive tutorials in which students can not only run pre-written code, but can also add their own code -- all with the aid of supporting explanatory text. All of the blog posts upon which EwP is based were written in Jupyter notebooks.

@@

The stage is set with an overview of a typical spreadsheet based model that most of the students have seen before.

@@

The notebook also includes a screenshot (Figure \ref{fig:xl_model}) of a typical Excel model.

@@

\subsubsection{Build first model - procedural approach}

We start with a very simple model that ignores the uncertainty in demand and that is not object oriented. Proceeding much like we would in Excel, base input values are stored in variables. Students are given skeleton code as scaffolding and must finish the expressions for computing key intermediate outputs (\code{order\_cost, sales\_revenue, refund\_revenue}) and the final output, \code{profit} -- see Figure \ref{fig:model_inputs}

\begin{figure}[!htbp]
	\centering
	\includegraphics[scale=0.5]{images/model_inputs.png}
	\caption{Instantiating model inputs}
	\label{fig:model_inputs}
\end{figure}

@@

At this point we discuss a fundamental difference between the computing model of Excel and that of Jupyter notebooks -- while spreadsheets respond automatically to changes, Jupyter notebooks require rerunning all code cells that include or follow the changed code. This difference in behavior between Jupyter notebooks and spreadsheets might be one of the most important concepts for students to internalize. A well known challenge, often referred to as the \textit{hidden state problem}, associated with computational notebooks is the ease of running code cells out of order and introducing inadvertent errors \cite{johnson2020benefits,grusDonNotebooks2018}. This point along with strategies to avoid it are discussed during the first week of the course which focuses on \href{http://www.sba.oakland.edu/faculty/isken/courses/aap/jupyter_notebooks.html}{notebook computing best practices as well as a historical look} at computational notebooks.

@@

Since order quantity is the key decision variable, we might want to see how profit changes for different order quantities. In Excel, the Data Table tool provides an easy way to do this and is a staple of spreadsheet based management science textbooks. In a Data Table, a range of order quantities can be used as row or column input and one or more output values can be computed. In Python, the vectorized nature of the \href{https://numpy.org/}{NumPy} library \cite{harris2020array} provides a simple way to accomplish the same thing. A range of order quantities is created as a NumPy array that can be used directly to compute vectors of all of the intermediate and final output variables. Much like Data Tables, vectorized computations avoid having to explicitly iterate, or loop, through a collection of input values.

A natural next step is to do the equivalent of an Excel 2-way Data Table. By creating a profit function that takes all of the base inputs as arguments, a list comprehension can then be used to generate the equivalent of a 2-way Data Table. Even better, this actually allows us to do the equivalent of an n-way Data Table with an arbitrary number of outputs -- something Excel does not have. This example provides an effective way to show the power of creating reusable functions and Python's handy list comprehension construct. As shown in Figure \ref{fig:function_datatable}, a few code and markdown cells are used to create the code, provide some explanation and show the output. This is the same output one would get with an Excel 2-way Data Table, just laid out differently. Then in Figure \ref{fig:datatable_scatter} we show how the output can be turned into a scatter plot with profit mapped to a color hue.

@@

Moving on to multiple uncertain inputs raises challenges much like those we faced when moving from a 1-way to an n-way data table function. In addition to modeling multiple uncertain inputs, we want to be able to specify a range of scenarios to run, much like we did for the \code{data\_table} function. The following set of design criteria are presented for a \code{simulate} function.

@@

This is followed by a well commented first attempt at a \code{simulate} function. By this point, students are well equipped to understand the code as it builds on basic Python concepts and more advanced concepts that we have already covered in the EwP module. The rest of the notebook is spent trying out the \code{simulate()} function and ends with creating a grouped boxplot, see Figure \ref{fig:boxplot} that many of them have likely created before with @Risk along with a faceted plot of histograms of profit created with the popular \href{https://seaborn.pydata.org/}{Seaborn} package.

@@

Now that we have created some useful modeling functionality, our goal is to make it easier for a modeler to actually use it. Certainly we do not want to rely on copying, pasting and modifying code when building a new model. In the Python world, the natural next step is to deploy our code as a Python \textit{package} \cite{buezenWelcomePythonPackages2022} that can then be imported like any other Python library.

@@

Key software design changes are then made which will make our code easier on our users. These changes include creating an abstract \code{Model} base class from which different model classes can be created and moving the \code{data\_table}, \code{goal\_seek}, and \code{simulate} functions into the \code{Model} base class as methods. All of the code is moved into a code module named \code{whatif.py}. Students are now ready to learn about the basics of turning our code into an installable package. This requires a basic understanding of two key Python files that are ubiquitous in Python pacakges - \code{\_\_init\_\_.py} and \code{setup.py}. Options are presented for installing our package locally so that we can use it for different modeling projects. This includes an explanation of how and where Python searches for packages when \code{import} statements are encountered in code. There is much complexity lurking here and students are reminded that they will need to revisit this topic frequently. I have made every effort to distill things down to a minimal level of complexity needed to convey the important points.

>>>>

\ABSTRACT{%
A staple of many spreadsheet based management science courses is the use of Excel for activities such as model building, sensitivity analysis, goal seeking and Monte-Carlo simulation. What might those things look like if carried out using Python? We describe a teaching module in which Python is used to do typical Excel based modeling and data wrangling tasks. In addition, students are exposed to basic software engineering principles including project folder structures, version control, object oriented programming and other more advanced Python skills, creating deployable packages and documentation. The module is supported with Jupyter notebooks, Python scripts, course web pages which include numerous screencasts, and a few GitHub repositories. All of the supporting materials are permissively licensed and freely accessible.
}%

@@

Spreadsheet based modeling has transformed the teaching of management science in business schools over the past twenty plus years. The late 1990's saw first editions of game changing textbooks \cite{winstonPracticalManagementScience2018,ragsdaleSpreadsheetModelingDecision2017} focused on the use of spreadsheets in teaching modeling and management science.  A series of conferences focused on teaching management science with spreadsheets started at Dartmouth in 1998 and several presentations from that conference made up the bulk of papers in the first two issues of the INFORMS Transactions on Education (ITE) journal. The next few decades saw numerous textbooks \cite{powellBusinessAnalyticsArt2019, camm2020business} and teaching papers in journals such as \href{https://pubsonline.informs.org/journal/ited}{ITE} that prominently featured spreadsheet based modeling.

Despite numerous calls for, and predictions of, the eradication of spreadsheets from the business world, Excel and other spreadsheets are \href{https://benn.substack.com/p/the-next-billion-programmers}{alive and well}. Yes, spreadsheets have well known limitations. Spreadsheets are not databases and probably should not be used as one, especially for relational data models. There are better data visualization and exploration tools. There are better dashboarding tools that are more easily maintained in a production environment. Spreadsheet errors can be a \href{https://eusprig.org/research-info/horror-stories/}{huge problem}. Version control is difficult as both data and logic live in the same spreadsheet document. Documentation of multi-step manual data manipulation processes is often lacking -- hindering reproducibility and automation. There are data size limitations and external data links present their \href{https://www.thespreadsheetguru.com/find-remove-external-links/}{own set of challenges}. But, Excel continues to be used for a wide range of quantitative analysis related activities throughout the business world. One particular use is the building and exercising of models of business problems
to explore the impact of important model inputs on different output metrics. This ``what if?'' type modeling is facilitated by Excel's flexibility, numerous built in functions, integrated sensitivity analysis via \href{https://support.microsoft.com/en-us/office/calculate-multiple-results-by-using-a-data-table-e95e2487-6ca6-4413-ad12-77542a5ea50b}{Data Tables},  \href{https://support.microsoft.com/en-us/office/use-goal-seek-to-find-the-result-you-want-by-adjusting-an-input-value-320cb99e-f4a4-417f-b1c3-4f369d6e66c7}{Goal
	Seek} for break-even analysis and exploratory data analysis via plots and Pivot Tables.

@@

At our institution, The EwP module is part of a business course entitled  \href{http://www.sba.oakland.edu/faculty/isken/courses/aap}{Advanced Analytics with Python} (AAP). The students taking this course have already taken the PCDA course within which they learned fundamental Python programming within the context of data analytics over seven weeks. The PCDA course ends with intro to the \href{https://scikit-learn.org/stable/index.html}{scikit-learn library} for doing predictive modeling in Python and the AAP course begins where the PCDA course ends. The EwP module could also be used as part of a semester long Python based analytics course. Most AAP students have also already taken   \href{http://www.sba.oakland.edu/faculty/isken/courses/ba}{Business Analytics} (BA) which is a spreadsheet based introduction to business analytics. The BA and PCDA courses have historically been offered in face to face mode in a computer teaching lab. Since 2020, these two courses have been also offered as online, asynchronous courses. The newer AAP course has been offered since 2021 in an online asynchronous mode. All three courses have extensive course websites that are publicly accessible, including access to all of the video content and supporting files. These courses serve upper level undergraduate and graduate students. Many are from the School of Business Administration but they draw students from many programs across our campus.

@@

\begin{itemize}
	\item variables, arithmetic and boolean operators, basic data types,
	\item data structures such as lists, dictionaries, tuples, NumPy arrays and pandas dataframes,
	\item flow control such as branching with \texttt{if \ldots else} and iterating with \texttt{for} and \texttt{while}
	\item module imports,
	\item using built in functions and accessing methods and properties of objects,
	\item creating functions,
	\item basic use of NumPy and pandas for data wrangling and analysis,
	\item basic plotting with matplotlib.
\end{itemize}

@@

All of the Python modules in the PCDA course as well as the entire AAP course are taught using a combination of Jupyter notebooks, Python scripts, and videos that walk readers through the notebooks. Jupyter notebooks, part of \href{https://jupyter.org/}{Project Jupyter}, are an example of \textit{literate computing} \cite{perez2015project} and allow one to create rich interactive documents that support iterative exploration and development.

Central to Jupyter notebooks is a mix of \textit{code cells}, \textit{markdown cells} and \textit{output cells}. Code cells contain executable code while markdown cells contain plain text using \href{URL}{markdown} for styling text. For example, Figure \ref{fig:cells_edit_mode} shows two markdown cells separated by a code cell as they look when in editing mode. After running the cells, they appear as show in Figure \ref{fig:cells_after_running}. After executing a code cell, the output appears directly below it in an output cell. Mixing code and markdown cells makes it quite easy to create interactive tutorials in which students can not only run pre-written code, but can also add their own code -- all with the aid of supporting explanatory text. All of the blog posts upon which EwP is based were written in Jupyter notebooks.

@@

The stage is set with an overview of a typical spreadsheet based model that most of the students have seen before. The notebook also includes a screenshot (Figure \ref{fig:xl_model}) of this Excel model.

@@

@@

\subsubsection{Model building using a procedural programming approach}

We start with a very simple model that ignores the uncertainty in demand and does not use object oriented programming concepts. Proceeding much like we would in Excel, base input values are stored in variables. Students are given skeleton code as scaffolding and must finish the expressions for computing key intermediate outputs (\code{order\_cost, sales\_revenue, refund\_revenue}) and the final output, \code{profit} -- see Figure \ref{fig:model_inputs}.

\begin{figure}[!htbp]
	\centering
	\includegraphics[scale=0.5]{images/model_inputs.png}
	\caption{Instantiating model inputs and creating output formulas}
	\label{fig:model_inputs}
\end{figure}

@@

At this point we discuss a fundamental difference between the computing model of Excel and that of Jupyter notebooks. While spreadsheets respond automatically to changes, Jupyter notebooks require rerunning all code cells that include or follow the changed code. This difference in behavior between Jupyter notebooks and spreadsheets might be one of the most important concepts for students to internalize. A well known challenge, often referred to as the \textit{hidden state problem}, associated with computational notebooks is the ease of running code cells out of order and introducing inadvertent errors \cite{johnson2020benefits,grusDonNotebooks2018}. This point along with strategies to avoid it are discussed during the first week of the course which focuses on \href{http://www.sba.oakland.edu/faculty/isken/courses/aap/jupyter_notebooks.html}{notebook computing best practices as well as a historical look} at computational notebooks.

@@

Since order quantity is the key decision variable, we might want to see how profit changes for different order quantities. In Excel, the Data Table tool provides a familiar and easy way to do this. In a Data Table, a range of order quantities can be used as row or column input and one or more output values can be computed. In Python, the vectorized nature of the \href{https://numpy.org/}{NumPy} library \cite{harris2020array} provides a simple way to accomplish the same thing. A range of order quantities is created as a NumPy array that can be used directly to compute vectors of all of the intermediate and final output variables. Much like Data Tables, vectorized computations avoid having to explicitly iterate, or loop, through a collection of input values.

A natural next step is to do the equivalent of an Excel 2-way Data Table. By creating a profit function that takes all of the base inputs as arguments, a \href{https://realpython.com/list-comprehension-python/}{list comprehension} can then be used to generate the equivalent of a 2-way Data Table. Even better, this actually allows us to do the equivalent of an n-way Data Table with an arbitrary number of outputs -- something Excel does not have. This example provides an effective way to show the power of creating reusable functions and Python's handy list comprehension construct. As shown in Figure \ref{fig:function_datatable}, a few code and markdown cells are used to create the code, provide some explanation and show the output. This is the same output one would get with an Excel 2-way Data Table, just laid out differently. Then in Figure \ref{fig:datatable_scatter} we show how the output can be turned into a scatter plot with profit mapped to a color hue.

@@

Moving on to multiple uncertain inputs raises challenges similar to those we faced when moving from a 1-way to an n-way data table function. In addition to modeling multiple uncertain inputs, we want to be able to specify a range of scenarios to run, just as we did for the \code{data\_table} function. Figure \ref{fig:oo_sim_design_specs} shows the design specifications for our \code{simulate} function.

@@

This is followed by a well commented first attempt at a \code{simulate} function. By this point, students are well equipped to understand the code as it builds on basic Python concepts and more advanced concepts that we have already covered in the EwP module. The rest of the notebook is spent trying out the \code{simulate()} function and ends with creating a grouped boxplot, see Figure \ref{fig:boxplot}, along with a faceted plot of profit histograms created with the popular \href{https://seaborn.pydata.org/}{Seaborn} package.

@@

Now that we have created some useful modeling functionality, our goal is to make it easier for a modeler to actually use the code. Certainly we do not want to rely on copying, pasting and modifying code when building a new model. In the Python world, the natural next step is to deploy our code as a Python \textit{package} \cite{buezenWelcomePythonPackages2022} that can then be imported like any other Python library.

@@

Key software design changes are then made which will make using our code easier on our users. These changes include creating an abstract \code{Model} base class from which different model classes can be created and moving the \code{data\_table}, \code{goal\_seek}, and \code{simulate} functions into the \code{Model} base class as methods. All of the code is moved into a code module named \code{whatif.py}. Students are now ready to learn about the basics of turning our code into an installable package. This requires a basic understanding of two key Python files that are ubiquitous in Python pacakges - \code{\_\_init\_\_.py} and \code{setup.py}. Options are presented for installing our package locally so that we can use it for different modeling projects. This includes an explanation of how and where Python searches for packages when \code{import} statements are encountered in code. There is much complexity lurking here and students are reminded that they will need to revisit this topic frequently. I have made every effort to distill things down to a minimal level of complexity needed to convey the important points.
